---
title: "Session 02: Working with Rectangular Data"
format: 
    html: 
        code-fold: true
        code-copy: false
        toc: true
        number-sections: false
execute:
    echo: true
---

## Learning goals 

- explain objects, functions, vectors, and data frames
- inspect, filter, and summarize data
- distinguish base R and the tidyverse

## Working with data frames in base R

### Generating synthetic (fake) data

We will generate a synthetic dataset for demonstration. The information in the dataset does not describe real students. 


```{r - generate-synthetic-data, echo = FALSE}
set.seed(2847) #this makes random values reproducible.

students <- data.frame(
  id = 1:12,
  group = sample(c("A", "B"), size = 12, replace = TRUE),
  study_hours = round(rnorm(12, mean = 6, sd = 1.8), 1),
  attended_lab = sample(
    c(TRUE, FALSE),
    size = 12,
    replace = TRUE,
    prob = c(0.75, 0.25)
  )
)
# Keep study hours at zero or above
students$study_hours <- pmax(students$study_hours, 0)

# Create a score related to study time and lab attendance
students$score <- round(
  48 + 4.5 * students$study_hours +
    7 * students$attended_lab + rnorm(12, mean = 0, sd = 5)
)

# Keep scores between 0 and 100
students$score <- pmin(pmax(students$score, 0), 100)

# Add two missing scores
students$score[c(4, 10)] <- NA

students

```


### Inspect a data frame

A data frame is an object that stores structured data in rectangular format where columns are variables and rows are observations.

```{r - inspect-synthetic-data}

head(students, 10) # Returns the first n rows of a data frame
dim(students)  # Returns the dimension (rows X columns) of a data frame
names(students)  # Returns the list of column names  
str(students) # Returns information about variable types and content
summary(students) # Returns information about distributional characteristics of variables
```

### Select columns and rows

Use `$` to select one column. 


```{r - select-data}}

students$study_hours
mean(students$study_hours)

students[1:5, c("id", "study_hours")]
```

We can also use square brackets as `data[rows, columns]`.

### ## Missing values

`NA` means that a value is missing. Comparisons with a missing value usually return `NA`, so many functions need `na.rm = TRUE`.

```{r - missing-values}

is.na(students$score)
sum(is.na(students$score))

mean(students$score)
mean(students$score, na.rm = TRUE)
```

## Base R and the tidyverse

- Base R is installed with R. All built-in functions, objects, and methods are what we call base R.   
- The **tidyverse** is a collection of R packages developed for common data science tasks, such as importing, organizing, transforming, and visualizing data. These packages share a common design philosophy, grammar, and data structures, which makes them work together consistently. 
- Learning some base R helps you understand the language, but **tidyverse** functions often make multi-step data work easier to read.

Packages provide additional functions and tools. Some packages come with R, while others, such as the **tidyverse**, must be installed separately. We install a package once and load it into each R session when we want to use it.


```{r - install and load tidyverse}
#install.packages("tidyverse")
library("tidyverse")
```

### The pipe and data verbs

The native R pipe `|>` sends the result on its left to the next function. Read it as **then**.


```{r - first pipeline}

students |>
    filter(study_hours >= 6) |>
    select(id, group, study_hours, score)
```

This pipeline translates into: 

1. Start with `students` data frame
2. Then keep rows where `study_hours` variable is greater than or equal to 6
3. Then keep these four columns: `id`, `group`, `study_hours`, and `score`

The same task in base R would look like this: 

```{r - baser R vs tidyverse, output = FALSE}

students[
  students$study_hours >= 6,
  c("id", "group", "study_hours", "score")
]

```


### Create and summarize variables with dplyr

`mutate()` creates or changes columns. `group_by()` and `summarise()` calculate group-level results.

```{r - mutate-and-summarise}

students_with_minutes <- students |>
  mutate(study_minutes = study_hours * 60)

students_with_minutes |>
  group_by(group) |>
  summarise(
    number_of_students = n(),
    average_score = mean(score, na.rm = TRUE)
  )
```


### Data frames and tibbles

A **tibble** is the tidyverse version of a data frame. It stores rectangular data in the same basic way, but prints large datasets more compactly.

```{r}
#| label: create-tibble

students_tbl <- as_tibble(students)

students_tbl
```

Most `dplyr` functions work with both ordinary data frames and tibbles.

### Arrange rows

`arrange()` changes the order of rows. By default, it sorts from the smallest value to the largest.

```{r - arrange-rows}
students
students |>
  arrange(score)
```

Use `desc()` to reverse the order.

```{r - arrange-descending}

students |>
  arrange(desc(score)) |>
  select(id, group, score)
```

`arrange()` does not remove rows or change their values. It only changes their order.

### Find and count categories

`distinct()` shows the unique values of one or more variables. `count()` reports how many observations belong to each category.

```{r - distinct-and-count}

students |>
  distinct(group)

students |>
  count(group)
```

### Filter missing and non-missing values

Use `is.na()` inside `filter()` to keep rows with missing values. Add `!`, meaning **not**, to keep rows without missing values.

```{r - filter-missing-values}

is.na(students$score)

# Students whose score is missing
students |>
  filter(is.na(score))

# Students whose score is not missing
students |>
  filter(!is.na(score))
```

### Combine several operations

Each verb performs one task. The pipe lets us combine them into a sequence that can be read from top to bottom.

```{r - combined-dplyr-operations}

students |>
  filter(!is.na(score)) |>
  arrange(desc(score)) |>
  select(id, group, study_hours, score)
```

Read this as: start with `students`, **then** remove missing scores, **then** sort from the highest score to the lowest, and **then** keep four columns.

## The NYC flights data

We now apply the same ideas to a real data set. The `nycflights13` package contains information about flights departing New York City in 2013.

```{r - load-flights}
library(nycflights13)

flights_all <- nycflights13::flights

```

Use the help page to read the variable descriptions.

```{r - label: flights-help, eval = FALSE}

?nycflights13::flights

```

The full data set has more than 300,000 rows. We will use September flights for most examples.

```{r - september-flights}

flights_sep <- flights_all |>
  filter(month == 9)

nrow(flights_all)
nrow(flights_sep)
```

### Inspect the data

`glimpse()` shows column names, data types, and example values.

```{r - inspect-flights}

dim(flights_sep)
glimpse(flights_sep)
```

You can also inspect the data in Positron without changing it.

```{r - view-flights, eval = FALSE}

View(flights_sep)
```

### Types and classes

`typeof()` reports how an object is stored. `class()` reports how R should treat it.

```{r - flights-variable-types}

typeof(flights_sep$dep_time)
class(flights_sep$dep_time)

typeof(flights_sep$arr_delay)
class(flights_sep$arr_delay)

typeof(flights_sep$origin)
class(flights_sep$origin)
```


### Missing values

Count missing departure times in base R.

```{r - missing-flight-values}

sum(is.na(flights_sep$dep_time))
summary(flights_sep$dep_time)
```

Count missing values in every column.

```{r - missing-all-columns}

colSums(is.na(flights_sep))
```

Do not remove missing values automatically. First ask what they mean and whether the variables are needed for the current analysis. 

### Categorical variables

`origin` is stored as character data, but it represents categories. Base R and `dplyr` can both count its values.

```{r - categorical-counts}

table(flights_sep$origin)

flights_sep |>
  count(origin)
```

Factors are useful when categories have a fixed set or order.

```{r - factors}

origin_factor <- factor(flights_sep$origin)
class(origin_factor)
levels(origin_factor)
```

### Date-time variables

`time_hour` has a special class, so date-time functions can operate on it.

```{r - datetimes}

head(flights_sep$time_hour)
class(flights_sep$time_hour)

table(weekdays(flights_sep$time_hour))
```

The tidyverse package `lubridate` provides additional date-time functions. For example, `wday()` can return weekday labels.

```{r - weekdays-lubridate}

flights_sep |>
  count(weekday = wday(time_hour, label = TRUE))
```