---
title: "Session 01: Getting Started with R"
format: 
    html: 
        code-fold: true
        code-copy: false
        toc: true
        number-sections: false
execute:
    echo: true
---

## Learning goals 

- identify the main parts of the Positron interface
- run R code in a Quarto (`.qmd`) document and preview the result
- understand objects, functions, vectors, and data frames


## Positron and the project folder

Open a folder for this course in Positron. Keeping the Quarto file and any data files in one project folder makes file paths simpler and your work easier to reproduce.

Positron may arrange panes differently depending on your window size, but these are the main components:

| Component | What it is for |
|---|---|
| Editor | Write and save `.qmd` and `.R` files. |
| Console | Run R commands and see immediate output. |
| Variables | See objects currently stored in memory. |
| Data Explorer | Inspect a data frame in a spreadsheet-like view. |
| Plots | View plots produced by R. |
| Help | Read documentation for functions and packages. |
| Viewer | Preview rendered Quarto documents. |


Check that an **R interpreter** is active. Then open this file and use the **Run Cell** button above a code cell. Use **Preview** to render the entire document.

## What is Quarto?

A Quarto document combines three things: 

1. a YAML header between the `---` lines;
2. Markdown text, headings, lists, and links;
3. Executable code cells.

The example below is an R code cell. Run it, change the numbers, and run it again. 


```{r - addition-operation, include = TRUE}
2 + 3
```

You can also set cell options that way: 


```{r - addition-operation}
#| label: addition-operation
#| include: true

2 + 2
```

Common options include: 

| Option | Effect |
|---|---|
| `echo: false` | Run the code but hide the code in the rendered document. |
| `output: false` | Run the code but hide its output. |
| `include: false` | Run the code but show neither code nor output. |
| `eval: false` | Show the code but do not run it. |

## Programming in R

Programming means giving a computer clear instructions. A computer processor cannot execute R code directly. The R interpreter reads and evaluates the R code, using lower-level computer instructions to carry out the requested operations. This allows us to write understandable R code instead of machine code. 

A common pattern in programming with R (also applies to Python) is: 

1. generate or load data
2. store it in an object
3. transform the object with a function
4. inspect or communicate the result

### Objects and assignment

The assignment operator `<-` gives a value a name and it stored in computer's memory with that name. 


```{r - objects in R }
course_name <- "Data Science Tools"
number_of_students <- 24
class_average <- 78.5

course_name
number_of_students
class_average
```

### Vectors and vectorized operations

A **vector** contains values of the same basic type. The function `c()` combines values into a vector. 


```{r - vectors}
study_hours <- c(3, 5, 4, 8, 6)
study_hours

#study_hours * 60
study_minutes <- study_hours * 60 # R applies this operation to every item stored in the vector
study_minutes

study_hours >= 5 # This comparison returns TRUE or FALSE for each item 
```


We can use square brackets to select specific elements. The condition we specify within square brackets tells R which elements to select.


```{r - vector-selection}

study_hours[study_hours>=5] 

```

### Functions and arguments

A functions performs an operation. Values supplied inside parantheses are its arguments.


```{r - functions}

mean(study_hours) # this is a built-in function.  

sum(study_hours)/length(study_hours) # this is another way to perform the same task, but redundant given we already have the fuction above. 

round(mean(study_hours))
```


We can use `?function_name` or `help(function_name)` to open documentation of a function. 


```{r - function-documentation}

?mean
help(mean)
```


#### Exercise 1
::: {.callout title="Exercise 1"}

Create a vector named `temperatures`. It must contain the values `18`, `20`, `19`, `23`, and `17`. Then calculate its mean.

```{r - exercise-1}
mean(c(18, 20, 19, 23, 17))
```

:::
