---
title: "Session 04: Building more complex data visualizations"
format:
  html:
    code-fold: true
    code-copy: false
    toc: true
    number-sections: false
execute:
  echo: true
---

## Learning goals
- Building more complex visualizations
- Generating density curves for analyzing a numerical variable's distribution
- Visualizing compositional makeup of quantities
- Fitting trend line to scatter plots
- Saving and exporting plots to use them outside your coding environment

## Loading packages and data

We will continue using the `nycflights13` data. The package contains several related dataframes that can be combined.

```{r load-packages-and-data}
# install.packages("nycflights13") # Run this line if needed.
# install.packages("tidyverse")
# install.packages("lubridate")

library(nycflights13)
library(tidyverse)

flights_all <- nycflights13::flights

flights_sep <- flights_all |>
  filter(month == 9)

head(flights_sep)
```


## Building more complex visualizations

The following examples combine data preparation, statistical summaries, scales, and several graphical layers. The goal is to make each design decision explicit.

### Controlling the order of categories

By default, categories are often shown alphabetically. We can calculate flight counts and use `reorder()` to arrange destinations by frequency.

```{r ordered-destination-bars}
top_destinations <- flights_sep |>
  count(dest, sort = TRUE) |>
  slice_head(n = 12)

top_destinations |>
  ggplot(
    mapping = aes(
      x = reorder(dest, n),
      y = n
    )
  ) +
  geom_col(
    fill = "steelblue",
    width = 0.75
  ) +
  coord_flip() +
  labs(
    title = "Most common destinations in September",
    subtitle = "Twelve destinations with the largest numbers of flights",
    x = NULL,
    y = "Number of flights"
  ) +
  theme_minimal()
```

`reorder(dest, n)` orders the destination codes using the values in `n`. `coord_flip()` exchanges the horizontal and vertical axes.

### Adding text to bars

Text can be added as another graphical layer. The same variables used to draw the bars can also determine the labels.

```{r labelled-destination-bars}
top_destinations |>
  ggplot(
    mapping = aes(
      x = reorder(dest, n),
      y = n
    )
  ) +
  geom_col(
    fill = "steelblue4",
    width = 0.75
  ) +
  geom_text(
    mapping = aes(label = n),
    hjust = -0.15,
    size = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Most common destinations in September",
    x = NULL,
    y = "Number of flights"
  ) +
  theme_minimal()
```

The `expand` argument creates space beyond the longest bar so that its label is not cut off.

### Proportions instead of counts

Counts answer how many flights belong to each category. Proportions answer what share of flights belongs to each category.

```{r origin-proportions-data}
origin_shares <- flights_sep |>
  count(origin) |>
  mutate(
    proportion = n / sum(n),
    percent = 100 * proportion
  )

origin_shares
```

```{r origin-proportions-plot}
origin_shares |>
  ggplot(
    mapping = aes(
      x = origin,
      y = proportion,
      fill = origin
    )
  ) +
  geom_col(
    width = 0.7,
    show.legend = FALSE
  ) +
  geom_text(
    mapping = aes(
      label = scales::percent(
        proportion,
        accuracy = 0.1
      )
    ),
    vjust = -0.4
  ) +
  scale_y_continuous(
    labels = scales::label_percent(),
    limits = c(0, 0.5)
  ) +
  labs(
    title = "Share of September flights by origin",
    x = "Origin airport",
    y = "Share of flights"
  ) +
  theme_minimal()
```

### Stacked bars

A stacked bar chart can show a total and its composition. Here, each bar represents a destination and each colored segment represents an origin airport.

```{r stacked-bars-data}
selected_destinations <- c(
  "ATL",
  "BOS",
  "LAX",
  "MIA",
  "ORD"
)

destination_origin_counts <- flights_sep |>
  filter(dest %in% selected_destinations) |>
  count(dest, origin)

destination_origin_counts
```

```{r stacked-destination-bars}
destination_origin_counts |>
  ggplot(
    mapping = aes(
      x = dest,
      y = n,
      fill = origin
    )
  ) +
  geom_col() +
  labs(
    title = "Flights to selected destinations",
    subtitle = "Bars are divided by origin airport",
    x = "Destination",
    y = "Number of flights",
    fill = "Origin"
  ) +
  theme_minimal()
```

### Filled bars for conditional proportions

Setting `position = "fill"` makes every bar the same height. This emphasizes composition rather than totals.

```{r proportional-stacked-bars}
destination_origin_counts |>
  ggplot(
    mapping = aes(
      x = dest,
      y = n,
      fill = origin
    )
  ) +
  geom_col(
    position = "fill"
  ) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  labs(
    title = "Origin composition by destination",
    x = "Destination",
    y = "Share of flights",
    fill = "Origin"
  ) +
  theme_minimal()
```

### Side-by-side bars

Setting `position = "dodge"` places groups beside each other, making absolute counts easier to compare.

```{r dodged-destination-bars}
destination_origin_counts |>
  ggplot(
    mapping = aes(
      x = dest,
      y = n,
      fill = origin
    )
  ) +
  geom_col(
    position = "dodge"
  ) +
  labs(
    title = "Flight counts by destination and origin",
    x = "Destination",
    y = "Number of flights",
    fill = "Origin"
  ) +
  theme_minimal()
```

### Comparing distributions with density curves

A density curve is a smoothed representation of a numerical distribution. It is useful for comparing the shapes of distributions.

```{r delay-density-curves}
flights_sep |>
  filter(
    !is.na(dep_delay),
    between(dep_delay, -30, 180)
  ) |>
  ggplot(
    mapping = aes(
      x = dep_delay,
      color = origin,
      fill = origin
    )
  ) +
  geom_density(
    alpha = 0.18,
    linewidth = 1
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey30"
  ) +
  labs(
    title = "Distribution of departure delays",
    subtitle = "Displayed range: -30 to 180 minutes",
    x = "Departure delay (minutes)",
    y = "Density",
    color = "Origin",
    fill = "Origin"
  ) +
  theme_minimal()
```


### Violin plots and boxplots together

A violin plot shows the shape of a distribution. A boxplot summarizes its median and quartiles. Layers allow us to display both.

```{r violin-and-boxplot}
flights_sep |>
  filter(
    !is.na(dep_delay),
    carrier %in% c("AA", "B6", "DL", "UA"),
    between(dep_delay, -30, 120)
  ) |>
  ggplot(
    mapping = aes(
      x = carrier,
      y = dep_delay,
      fill = carrier
    )
  ) +
  geom_violin(
    alpha = 0.55,
    color = NA,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.16,
    outlier.shape = NA,
    fill = "white"
  ) +
  guides(fill = "none") +
  labs(
    title = "Departure-delay distributions",
    subtitle = "Violin shapes with boxplot summaries",
    x = "Carrier",
    y = "Departure delay (minutes)"
  ) +
  theme_minimal()
```


### Adding a fitted trend

`geom_smooth()` estimates a relationship between two numerical variables. Here we compare the points with a linear trend.

```{r scatterplot-with-linear-trend}
set.seed(2026)

flights_sep |>
  filter(
    !is.na(air_time),
    !is.na(distance)
  ) |>
  slice_sample(n = 1000) |>
  ggplot(
    mapping = aes(
      x = air_time,
      y = distance
    )
  ) +
  geom_point(
    alpha = 0.15,
    color = "steelblue"
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "firebrick",
    fill = "mistyrose"
  ) +
  labs(
    title = "Distance and time in the air",
    subtitle = "Points with an estimated linear trend",
    x = "Time in the air (minutes)",
    y = "Distance (miles)"
  ) +
  theme_minimal()
```


Let's generate some synthetic data to explore different relationship patterns. 

```{r synthetic-data-generation}
set.seed(2998)

x = rnorm(1000, 50, 10)
y = 1.5*x + rnorm(1000,40,10) # we are generating y as a function of x 

fake_data = tibble(x = x, y = y)

fake_data |> 
  glimpse()
```

We will now explore the direction and strength of this relationship with a scatter plot. 

```{r scatterplot-with-trend-synthetic-data}
set.seed(2026)

fake_data |>
  ggplot(
    mapping = aes(
      x = x,
      y = y
    )
  ) +
  geom_point(
    alpha = 0.15,
    color = "steelblue"
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "firebrick",
    fill = "mistyrose"
  ) +
  labs(
    title = "Relationship between x and y"
  ) +
  theme_minimal()
```

Try to see how the data and the fitted line changes when you change x variable's coefficient in y variable equation. 

### Visualizing missing values over time

Missingness can itself be informative. We can calculate the daily percentage of flights with missing departure-delay values.

```{r daily-missingness-data}
daily_missingness <- flights_sep |>
  mutate(
    date = as.Date(
      paste(year, month, day, sep = "-")
    )
  ) |>
  group_by(date, origin) |>
  summarize(
    total_flights = n(),
    missing_dep_delay = sum(
      is.na(dep_delay)
    ),
    missing_share = mean(
      is.na(dep_delay)
    ),
    .groups = "drop"
  )

daily_missingness
```

```{r daily-missingness-plot}
daily_missingness |>
  ggplot(
    mapping = aes(
      x = date,
      y = missing_share,
      color = origin
    )
  ) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 1.5
  ) +
  scale_y_continuous(
    labels = scales::label_percent(
      accuracy = 0.1
    )
  ) +
  scale_x_date(
    date_breaks = "1 week",
    date_labels = "%b %d"
  ) +
  labs(
    title = "Missing departure-delay values by day",
    x = NULL,
    y = "Flights with missing departure delay",
    color = "Origin"
  ) +
  theme_minimal()
```

### Saving a plot as an object

A plot can be stored in an R object. This makes it easier to print, modify, or save later.

```{r create-plot-object}
distance_time_plot <- flights_sep |>
  filter(
    !is.na(air_time),
    !is.na(distance)
  ) |>
  slice_head(n = 1500) |>
  ggplot(
    mapping = aes(
      x = air_time,
      y = distance,
      color = origin
    )
  ) +
  geom_point(
    alpha = 0.3
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Distance and time in the air",
    x = "Time in the air (minutes)",
    y = "Distance (miles)",
    color = "Origin"
  ) +
  theme_minimal()

distance_time_plot
```

### Saving a plot to a file

`ggsave()` saves the most recently displayed plot or a named plot object. 

```{r save-plot-file, eval=FALSE}
ggsave(
  filename = "distance-time-plot.png",
  plot = distance_time_plot,
  width = 8,
  height = 5,
  units = "in",
  dpi = 300
)
```

