---
title: "Session 05: Working with relational data and time series data"
format:
  html:
    code-fold: true
    code-copy: false
    toc: true
    number-sections: false
execute:
  echo: true
---

## Learning goals

- Understanding the unit of observation and identifying keys
- Combining related dataframes with joins
- Working with dates, times, strings, and factors
- Constructing and auditing an analysis-ready dataset

## Loading packages and data

We will continue using the `nycflights13` data. This package contains several related dataframes. So far, we only used `flights` dataset, which contains information at flight level. 

```{r load-packages-and-data, output = FALSE}
# install.packages("nycflights13") # Run this line if needed.
# install.packages("tidyverse")
# install.packages("lubridate")

library(nycflights13)
library(tidyverse)
library(lubridate)

flights_all <- nycflights13::flights

set.seed(2026)

flights_sample <- flights_all |>
  slice_sample(n = 10000) |>
  mutate(sample_id = row_number())

head(flights_sample)
```

`set.seed()` makes the random draw reproducible. `sample_id` is a simple identifier for rows in this sample; a flight number alone does not uniquely identify a flight.

The package contains five main dataframes:

- `flights`: individual flights departing New York City
- `airlines`: airline names and carrier codes
- `airports`: airport names and locations
- `planes`: information about aircraft
- `weather`: hourly weather measurements

```{r inspect-related-dataframes}
#glimpse(airlines)
#glimpse(airports)
glimpse(planes)
#glimpse(weather)
```


## The unit of observation

The **unit of observation** is what one row represents.

In `flights_sample`, one row represents one scheduled flight.

```{r unit-of-observation-flights}
flights_sample |>
  select(year, month, day, carrier, flight, origin, dest) |>
  slice_head(n =10)
```

In `airlines`, one row represents one airline.

```{r unit-of-observation-airlines}
airlines
```

In `airports`, one row represents one airport.

```{r unit-of-observation-airports}
airports |>
  slice_head(n = 10)
```

The unit of observation determines what questions the dataframe can answer. It also determines how we should join, summarize, and model the data. 


## Identifiers and keys

A **key** is one or more variables that identify observations or connect dataframes. The carrier code is unique in `airlines`.

```{r check-airline-key}
airlines <- nycflights13::airlines
glimpse(airlines)

length(unique(airlines$carrier))
length(unique(airlines$carrier)) == nrow(airlines) # base R way to check
 

airlines |>
  count(carrier) |>
  filter(n > 1)

```

An empty result indicates no repeated carrier codes. For the weather data, the useful key is **both** `origin` and `time_hour`.

```{r check-weather-key}
weather <- nycflights13::weather

weather |>
  slice_head()

repeated_rows <- weather |>
  select(origin, year, month, day, hour, time_hour) |>
  add_count(origin, year, month, day, hour) |>
  filter(n > 1)
  

repeated_rows

as.numeric(repeated_rows$time_hour)

weather |>
  count(origin, time_hour) |>
  filter(n > 1)
```

This check should also return no rows. By contrast, `origin`, `year`, `month`, `day`, and `hour` do not uniquely identify every weather row: the 1 a.m. hour occurs twice when daylight saving time ends. The date-time value `time_hour` distinguishes those observations.


## Joining flights and weather

Both `flights` and `weather` contain `time_hour`. In `flights`, it is the scheduled departure hour; in `weather`, it is the observation hour. We match on this date-time and the origin airport. We select only the needed weather columns before joining.

```{r prepare-weather-lookup}
weather_lookup <- weather |>
  select(
    origin,
    time_hour,
    temp,
    wind_speed,
    precip
  ) |>
  mutate(weather_recorded = TRUE)

glimpse(weather_lookup)
```

Each sampled flight can match at most one weather observation. Many flights may share one airport-hour, so the expected relationship is many-to-one.

```{r join-weather-to-flights}
flights_weather <- flights_sample |>
  left_join(
    weather_lookup,
    by = join_by(origin, time_hour),
    relationship = "many-to-one"
  ) |>
  mutate(
    weather_recorded = replace_na(weather_recorded, FALSE)
  )

flights_weather |>
  select(
    sample_id,
    origin,
    sched_dep_time,
    time_hour,
    dep_delay,
    temp,
    wind_speed,
    precip
  ) |>
  slice_head(n = 10)
```

`left_join()` keeps every sampled flight. An unmatched flight gets `NA` in the added weather columns, while `weather_recorded` becomes `FALSE`.

```{r audit-weather-join}
nrow(flights_sample)
nrow(flights_weather)
nrow(flights_sample) == nrow(flights_weather)
n_distinct(flights_weather$sample_id) == nrow(flights_weather)

flights_weather |>
  summarize(
    no_weather_match = sum(!weather_recorded),
    matched_but_temp_missing = sum(
      weather_recorded & is.na(temp)
    )
  )
```

These last two counts answer different questions. A matched weather row may still have a missing temperature measurement.

`anti_join()` shows flights with no matching airport-hour in the weather dataset.

```{r inspect-unmatched-weather-keys}
flights_sample |>
  anti_join(
    weather_lookup,
    by = join_by(origin, time_hour)
  ) |>
  select(sample_id, origin, time_hour) |>
  slice_head(n = 10)
```

### Adding airline names

Airline operations may also relate to departure delays. We can add names without changing the flight-level unit of observation.

```{r join-airline-names}
flights_weather <- flights_weather |>
  left_join(
    airlines,
    by = join_by(carrier),
    relationship = "many-to-one"
  ) |>
  rename(airline_name = name)

nrow(flights_weather)

flights_weather |>
  select(carrier, airline_name, origin, dep_delay) |>
  slice_head(n = 10)
```

Other joins answer different questions: `inner_join()` keeps matches on both sides; `semi_join()` keeps left-hand rows with a match; `anti_join()` keeps left-hand rows without one.

## Dates, times, strings, and factors

The `time_hour` column is a `POSIXct` date-time. The year, month, and day columns allow us to make a calendar date without losing the flight's recorded departure date.

```{r create-time-and-label-variables}
flights_weather <- flights_weather |>
  mutate(
    flight_date = make_date(year, month, day),
    weekday = wday(
      flight_date,
      label = TRUE,
      week_start = 1
    ),
    week_start = floor_date(
      flight_date,
      unit = "week",
      week_start = 1
    ),
    scheduled_hour = hour,
    flight_code = str_c(carrier, flight, sep = "-")
  )

flights_weather |>
  select(
    flight_code,
    flight_date,
    weekday,
    time_hour,
    week_start,
    scheduled_hour
  ) |>
  slice_head(n = 10)
```

`str_c()` joins pieces of text into a readable flight label. `wday(label = TRUE)` creates an ordered factor of weekday names. A factor represents categories with an explicit order, useful for tables and plots. `week_start` groups dates into calendar weeks; the first week may begin in December 2012 even though every sampled flight occurred in 2013.

## Creating variables for the weather question

We define rain from precipitation, stronger wind from wind speed, and a 30-minute departure-delay indicator. Missing measurements remain missing.

```{r create-analysis-variables}
flights_weather <- flights_weather |>
  mutate(
    rainy = precip > 0,
    strong_wind = wind_speed >= 20,
    delayed_30 = dep_delay >= 30,
    weather_condition = case_when(
      !weather_recorded ~ "No weather match",
      is.na(precip) | is.na(wind_speed) ~ "Incomplete weather",
      rainy & strong_wind ~ "Rain and stronger wind",
      rainy ~ "Rain",
      strong_wind ~ "Stronger wind",
      TRUE ~ "No rain / lighter wind"
    ),
    weather_condition = factor(
      weather_condition,
      levels = c(
        "No rain / lighter wind",
        "Rain",
        "Stronger wind",
        "Rain and stronger wind",
        "Incomplete weather",
        "No weather match"
      )
    )
  )

flights_weather |>
  count(weather_condition, .drop = FALSE)
```

The 20 mph wind threshold is a teaching choice. A different threshold could change the comparison, so it must be reported. `case_when()` evaluates its conditions from top to bottom. The factor levels keep the categories in a deliberate order.

## Basic exploration of delays and weather

First, inspect missing outcomes and weather coverage. A missing departure delay must not be replaced with zero.

```{r missingness-audit}
flights_weather |>
  summarize(
    sampled_flights = n(),
    missing_dep_delay = sum(is.na(dep_delay)),
    no_weather_match = sum(!weather_recorded),
    missing_precip = sum(is.na(precip)),
    missing_wind_speed = sum(is.na(wind_speed))
  )
```

For a direct rain comparison, use flights with a known departure delay and precipitation measurement. This defines the denominator for both groups.

```{r summarize-delays-by-rain}
rain_comparison <- flights_weather |>
  filter(
    !is.na(dep_delay),
    !is.na(rainy)
  ) |>
  group_by(rainy) |>
  summarize(
    flights = n(),
    mean_delay = mean(dep_delay),
    median_delay = median(dep_delay),
    share_delayed_30 = mean(delayed_30),
    .groups = "drop"
  ) |>
  mutate(
    rain_status = if_else(rainy, "Rain", "No rain")
  )

rain_comparison
```

The median is less sensitive than the mean to a few very large delays. The groups may also differ in airport, season, time of day, or airline; this table alone cannot isolate an effect of rain.

```{r plot-rain-comparison}
rain_comparison |>
  ggplot(
    mapping = aes(
      x = rain_status,
      y = mean_delay
    )
  ) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Mean departure delay by scheduled-hour rain",
    subtitle = "Flights with observed delay and precipitation",
    x = NULL,
    y = "Mean departure delay (minutes)"
  ) +
  theme_minimal()
```

We can inspect the wind relationship without deciding immediately on a single threshold.

```{r explore-wind-bands}
flights_weather |>
  filter(
    !is.na(dep_delay),
    !is.na(wind_speed)
  ) |>
  mutate(
    wind_band = cut(
      wind_speed,
      breaks = c(-Inf, 5, 10, 15, 20, Inf),
      right = FALSE
    )
  ) |>
  group_by(wind_band) |>
  summarize(
    flights = n(),
    mean_delay = mean(dep_delay),
    .groups = "drop"
  )
```

Check group sizes before interpreting any average based on only a few flights.

## Reshaping a weather comparison

We can summarize rain and no-rain flights within each origin airport. Each row below is an airport and rain-status combination.

```{r origin-rain-long}
origin_rain_long <- flights_weather |>
  filter(
    !is.na(dep_delay),
    !is.na(rainy)
  ) |>
  mutate(
    rain_status = if_else(rainy, "rain", "no_rain")
  ) |>
  group_by(origin, rain_status) |>
  summarize(
    mean_delay = mean(dep_delay),
    flights = n(),
    .groups = "drop"
  )

origin_rain_long
```

`pivot_wider()` places the two rain-status means in separate columns. Each row then represents one airport.

```{r origin-rain-wide}
origin_rain_wide <- origin_rain_long |>
  select(origin, rain_status, mean_delay) |>
  pivot_wider(
    names_from = rain_status,
    values_from = mean_delay
  )

origin_rain_wide
```

`pivot_longer()` reverses that layout. It restores one row per airport and rain status.

```{r origin-rain-long-again}
origin_rain_wide |>
  pivot_longer(
    cols = c(no_rain, rain),
    names_to = "rain_status",
    values_to = "mean_delay"
  )
```

Reshaping changes the unit of observation of the summary table. It does not alter the flight-level dataset.
