---
title: "Session 03: Data visualization with ggplot"
format: 
    html: 
        code-fold: true
        code-copy: false
        toc: true
        number-sections: false
execute:
    echo: true
---

## Learning goals 

- Learning the basic plot types and implementing them with ggplot2
- Learning the logic of mapping variables to visual properties
- Building more complex visualizations with graphical layers

## Loading packages and data

We will again use data from the `nycflights13` package. We need to load the package before generating any visualizations. 

```{r - load-packages-and-data, output = FALSE}
#install.packages("nycflights13") # run this line if you do not have it installed in your coding environment. 
#install.packages("tidyverse")

library(nycflights13)
library(tidyverse)

flights_all <- nycflights13::flights # We will again mostly use a subset of the full dataset which only contains September data. 

flights_sep <- flights_all |> 
  filter(month == 9)
head(flights_sep)
```

## Introduction to data visualization with ggplot2

The `ggplot2` package creates visualizations by combining data, variable mappings, and graphical layers.

A basic plot usually contains three parts:

- `ggplot()` specifies the data.
- `aes()` maps variables to visual properties such as position, color, shape, or size.
- A `geom_*()` function determines how observations are displayed.

The `+` operator adds layers to a plot.

### A first plot

A scatter plot shows the relationship between two numerical variables. Each point represents one flight.

```{r first-ggplot}
flights_sep |>
  filter(!is.na(air_time), !is.na(distance)) |>
  slice_head(n = 1000) |>
  ggplot(
    mapping = aes(x = air_time, y = distance)
  ) +
  geom_point()
```

The pipe passes the prepared data to `ggplot()`. Therefore, we do not need to write `data = flights_sep` inside `ggplot()`.

Inside `aes()`:

- `air_time` determines the horizontal position.
- `distance` determines the vertical position.

### Fixed and mapped aesthetics

An **aesthetic** is a visual property of a plot, such as position, color, shape, or size.

A value written outside `aes()` applies to every observation.

```{r fixed-point-color}
flights_sep |>
  filter(!is.na(air_time), !is.na(distance)) |>
  slice_head(n = 1000) |>
  ggplot(
    mapping = aes(x = air_time, y = distance)
  ) +
  geom_point(
    color = "magenta4",
    alpha = 0.5
  )
```

Here, every point has the same color. The `alpha` argument controls transparency.

R has 657 built in colors. We can view the list of built-in colors. 

```{r - built-in-colors}
?ggplot2::geom_point
grDevices::colors()[1:10]
```

A variable written inside `aes()` is mapped to an aesthetic.

```{r mapped-point-color}
flights_sep |>
  filter(!is.na(air_time), !is.na(distance)) |>
  slice_head(n = 1000) |>
  ggplot(
    mapping = aes(
      x = air_time,
      y = distance,
      color = origin
    )
  ) +
  geom_point(alpha = 0.5)
```

Here, color represents the flight's airport of origin. Because `origin` is mapped to color, `ggplot2` automatically creates a legend.

### Bar charts for categorical variables

A bar chart shows how many observations belong to each category.

We will now examine the number of flights for several selected destinations. Let's first see what are all the destinations available at NYC airports.

```{r - list-all-destinations}
flights_sep |> distinct(dest)
```

Visualizing data for all destinations will make the plot look too cluttered. Looking at a few major ones will work for now. 

```{r destination-bar-chart}
flights_sep |>
  filter(dest %in% c("ATL", "BOS", "LAX", "MIA", "ORD")) |>
  ggplot(
    mapping = aes(x = dest)
  ) +
  geom_bar()
```

`geom_bar()` counts the observations in each category automatically. Its y-axis displays the number of flights.

### `geom_bar()` and `geom_col()`

Use `geom_bar()` when `ggplot2` should count the observations. Use `geom_col()` when the data already contain the values that determine the heights of the bars.

First, we can calculate the counts ourselves.

```{r count-destinations}
destination_counts <- flights_sep |>
  filter(dest %in% c("ATL", "BOS", "LAX", "MIA", "ORD")) |>
  count(dest)

destination_counts
```

The variable `n` contains the number of flights to each destination.

```{r destination-column-chart}
destination_counts |>
  ggplot(
    mapping = aes(x = dest, y = n)
  ) +
  geom_col()
```

These two bar charts display the same information, but the counts are calculated at different stages.

### Histograms for numerical variables

A histogram displays the distribution of a numerical variable. It divides the values into intervals called **bins**.

```{r air-time-histogram}
flights_sep |>
  filter(!is.na(air_time)) |>
  ggplot(
    mapping = aes(x = air_time)
  ) +
  geom_histogram(
    binwidth = 30,
    fill = "steelblue",
    color = "white"
  )
```

The `binwidth` argument controls the width of the intervals. Here, each interval covers 30 minutes of flight time.

Different bin widths can reveal different features of the distribution.

```{r air-time-larger-binwidth}
flights_sep |>
  filter(!is.na(air_time)) |>
  ggplot(
    mapping = aes(x = air_time)
  ) +
  geom_histogram(
    binwidth = 100,
    fill = "steelblue",
    color = "white"
  )
```

It is useful to try several reasonable bin widths before choosing one.

::: {.callout-note collapse="false"} 

`binwidth` controls bin width, but not where the first bin begins. To control where the first bin begins on the x axis, we can use `boundary`. When this argument is not supplied within a `geom_` function, `ggplot2` determines first bin position automatically and places the remaining bins at intervals equal to the specified binwidth. Below is an example with `boundary`:

:::

```{r air-time-histogram-boundary}
flights_sep |>
  filter(!is.na(air_time)) |>
  ggplot(
    mapping = aes(x = air_time)
  ) +
  geom_histogram(
    binwidth = 100,
    boundary = 0,
    fill = "steelblue",
    color = "white"
  )
```


### Boxplots for numerical and categorical variables

A boxplot can compare the distribution of a numerical variable across categories.

```{r departure-delay-boxplot}
flights_sep |>
  filter(
    !is.na(dep_delay),
    carrier %in% c("AA", "B6", "DL", "UA")
  ) |>
  ggplot(
    mapping = aes(x = carrier, y = dep_delay)
  ) +
  geom_boxplot()
```

The line inside each box represents the median. The box contains the middle half of the observed values. Points beyond the whiskers are potential outliers.

Extreme delays make the boxes difficult to see. We can focus the displayed y-axis without removing observations from the calculation.

```{r departure-delay-boxplot-zoom}
flights_sep |>
  filter(
    !is.na(dep_delay),
    carrier %in% c("AA", "B6", "DL", "UA")
  ) |>
  ggplot(
    mapping = aes(x = carrier, y = dep_delay)
  ) +
  geom_boxplot() +
  coord_cartesian(ylim = c(-30, 120)) + 
  theme_minimal()
```

`coord_cartesian()` zooms in on part of the plot. It does not filter observations from the data.

### Adding labels and a theme

We can use `labs()` to add an informative title and clearer axis labels. A `theme` controls the plot's general appearance.

```{r labelled-flight-plot}
flights_sep |>
  filter(!is.na(air_time), !is.na(distance)) |>
  slice_head(n = 10000) |>
  ggplot(
    mapping = aes(
      x = air_time,
      y = distance,
      color = origin
    )
  ) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Flight distance and time in the air",
    x = "Time in the air (minutes)",
    y = "Distance (miles)",
    color = "Origin airport"
  ) +
  theme_minimal()
```

Good labels help readers understand a visualization without having to inspect the code.

### Small multiples with facets

Faceting divides a plot into panels. Each panel displays a subset of the data.

```{r flight-facets}
flights_sep |>
  filter(!is.na(air_time), !is.na(distance)) |>
  slice_head(n = 1000) |>
  ggplot(
    mapping = aes(x = air_time, y = distance)
  ) +
  geom_point(alpha = 0.4, color = "steelblue") +
  facet_wrap(~ origin) +
  labs(
    title = "Flight distance and time by origin airport",
    x = "Time in the air (minutes)",
    y = "Distance (miles)"
  ) +
theme_light()
```

`facet_wrap(~ origin)` creates one panel for each origin airport. Because the panels use the same axes, their patterns can be compared.